---
title: "Using teal.picks in a teal application"
author: "NEST CoreDev Team"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Using teal.picks in a teal application}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  rmarkdown::html_vignette:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = NA)
```

## What this package is for

**teal.picks** helps you expose **analysis choices** in a **teal** Shiny app in a way that matches how people think about data: first *which dataset*, then *which columns*, and optionally *which values* to filter to. You describe those steps once with **`picks()`**; the app renders the right controls and keeps selections consistent when upstream choices change.

Typical workflow:

1. Prepare a **[`teal.data::teal_data()`]** object (and, when you relate multiple tables, **join keys** so teal knows how they link).
2. Define one or more **`picks()`** specs: **`datasets()`**, optionally **`variables()`**, optionally **`values()`** for filters or ranges.
3. In your **teal module**, wire **UI and server** so those specs drive your tables, plots, or merged analysis data (see [Writing your own module](#writing-your-own-module)).

The examples below show the usual building blocks—not every app needs every level.

```{r data}
library(teal)
library(teal.data)
library(teal.picks)

data <- teal_data()
data <- within(data, {
  ADSL <- data.frame(
    USUBJID = sprintf("S%03d", 1:12),
    AGE = sample(40:75, 12, replace = TRUE),
    SEX = rep(c("M", "F"), 6),
    stringsAsFactors = FALSE
  )
  ADLB <- data.frame(
    USUBJID = rep(sprintf("S%03d", 1:12), each = 4),
    PARAM = rep(c("ALT", "AST", "CRP", "GLU"), 12),
    AVAL = round(rnorm(48, mean = 40, sd = 8), 1),
    stringsAsFactors = FALSE
  )
})

join_keys(data) <- join_keys(teal.data::join_key("ADSL", "ADLB", keys = "USUBJID"))
```

### Choose a dataset

When the analysis can draw from more than one table, let the user pick the active source (here, demographics vs labs).

```{r picks-datasets}
picks_datasets <- list(
  source = picks(
    datasets(choices = c("ADSL", "ADLB"), selected = "ADSL")
  )
)
```

### Choose dataset and columns

Add **`variables()`** so, after a dataset is chosen, the user picks one or more columns. You can list names explicitly or use **tidyselect**-style expressions; see **`?picks`** for details.

```{r picks-datasets-variables}
picks_datasets_variables <- list(
  adsl_cols = picks(
    datasets(choices = "ADSL", selected = "ADSL"),
    variables(
      choices = c("USUBJID", "AGE", "SEX"),
      selected = "AGE",
      multiple = FALSE
    )
  )
)
```

### Add a value filter (levels or ranges)

**`values()`** sits after **`variables()`**. It adapts to the column type—for example category levels for a **`PARAM`**-style variable, or a numeric or date range for continuous data.

```{r picks-datasets-variables-values}
picks_datasets_variables_values <- list(
  labs = picks(
    datasets(choices = "ADLB", selected = "ADLB"),
    variables(choices = "PARAM", selected = "PARAM", multiple = FALSE),
    values(
      choices = c("ALT", "AST", "CRP", "GLU"),
      selected = c("ALT", "AST"),
      multiple = TRUE
    )
  )
)
```

### Several columns at once

Use **`multiple = TRUE`** when analysts should pass more than one variable into the next step (for example stratifiers or outcomes together).

```{r picks-multiple-variables}
picks_multiple_variables <- list(
  demo = picks(
    datasets(choices = "ADSL", selected = "ADSL"),
    variables(
      choices = c("USUBJID", "AGE", "SEX"),
      selected = c("AGE", "SEX"),
      multiple = TRUE,
      ordered = TRUE
    )
  )
)
```

## Trying these patterns inside teal

You can paste the **`picks`** objects above into any module that consumes them. To **run** the four patterns without writing module code first, the package includes **`tm_merge()`**: a small **example** teal module that connects **`picks`** to a merged preview table. That is **not** required for normal use—it is simply a convenient host while you learn the **`picks()`** API.

Run the next block **interactively** on your machine (it is not executed during package checks).

```{r teal-app, eval = FALSE}
library(shiny)

app <- init(
  data = data,
  modules = modules(
    modules(
      label = "teal.picks patterns",
      tm_merge(
        label = "1. Dataset choice",
        picks = picks_datasets
      ),
      tm_merge(
        label = "2. Dataset & variables",
        picks = picks_datasets_variables
      ),
      tm_merge(
        label = "3. Dataset, variables & values",
        picks = picks_datasets_variables_values
      ),
      tm_merge(
        label = "4. Multiple variables",
        picks = picks_multiple_variables
      )
    )
  )
)

shinyApp(app$ui, app$server, enableBookmarking = "server")
```

## Writing your own module {#writing-your-own-module}

In real apps you usually integrate **`picks`** with your own **`teal::module()`**. Two functions do the wiring:

- **[`picks_ui()`]** — place the selector controls in your module UI (pass the same **`picks`** object you defined for that module).
- **[`picks_srv()`]** — in the module server, pass **`data`** as a reactive **`teal_data`** object; it returns the **resolved** picks (what the user chose, after any dynamic choices are applied).

Skeleton (not run here):

```{r module-skeleton, eval = FALSE}
# Define once before init(), then pass the same object via ui_args / server_args:
# my_picks <- picks(datasets(...), variables(...))  # optional: values(...)

teal::module(
  label = "My analysis",
  ui = function(id, picks) {
    ns <- shiny::NS(id)
    shiny::tags$div(
      teal.picks::picks_ui(ns("sel"), picks = picks),
      # ... outputs that use selections ...
    )
  },
  server = function(id, data, picks) {
    shiny::moduleServer(id, function(input, output, session) {
      selections <- teal.picks::picks_srv("sel", picks = picks, data = data)
      # selections() is reactive; use it to filter data or build ggplot/tables/etc.
    })
  },
  ui_args = list(picks = my_picks),
  server_args = list(picks = my_picks),
  datanames = "all"
)
```

See **`?picks_ui`** and **`?picks_srv`** for container options and multiple named selectors.

## See also

- **`?picks`** — full **`datasets()`** / **`variables()`** / **`values()`** reference.
- **`vignette("teal-picks-standalone-shiny", package = "teal.picks")`** — **`picks_ui()`** / **`picks_srv()`** in plain Shiny (no **`init()`**).
- **teal** documentation for **`init()`**, **`modules()`**, and **`teal::module()`**.
