---
title: "teal.picks without teal (standalone Shiny)"
author: "NEST CoreDev Team"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{teal.picks without teal (standalone Shiny)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  rmarkdown::html_vignette:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = NA)
```

## Introduction

You can use **`picks_ui()`** and **`picks_srv()`** in a plain **Shiny** app: pass a reactive [`teal.data::teal_data()`] object to **`picks_srv()`** and combine results with **`merge_srv()`** when you need merged analysis data. This mirrors what **`tm_merge()`** does inside **`teal`**, without **`teal::init()`**.

Run the **`shinyApp`** chunk **interactively**.

```{r data}
library(shiny)
library(teal.data)
library(teal.picks)

data <- teal_data()
data <- within(data, {
  ADSL <- data.frame(
    USUBJID = sprintf("S%03d", 1:8),
    AGE = sample(35:70, 8, replace = TRUE),
    stringsAsFactors = FALSE
  )
  ADLB <- data.frame(
    USUBJID = rep(sprintf("S%03d", 1:8), each = 3),
    PARAM = rep(c("ALT", "AST", "BILI"), 8),
    AVAL = round(rnorm(24, 42, 6), 1),
    stringsAsFactors = FALSE
  )
})

join_keys(data) <- join_keys(teal.data::join_key("ADSL", "ADLB", keys = "USUBJID"))

selector_def <- picks(
  datasets(choices = c("ADSL", "ADLB"), selected = "ADLB"),
  variables(
    choices = tidyselect::everything(),
    selected = c(1L, 2L),
    multiple = TRUE
  )
)
```

## Minimal Shiny app

```{r standalone-app, eval = FALSE}
ui <- fluidPage(
  titlePanel("Standalone picks + merge"),
  fluidRow(
    column(
      width = 4,
      picks_ui("sel", picks = selector_def)
    ),
    column(
      width = 8,
      tags$h4("Mapped variables"),
      verbatimTextOutput("mapped"),
      tags$h4("Merge preview"),
      tableOutput("merged")
    )
  )
)

server <- function(input, output, session) {
  data_r <- reactive(data)

  selectors <- list(sel = picks_srv("sel", picks = selector_def, data = data_r))

  merged <- merge_srv(
    id = "merge",
    data = data_r,
    selectors = selectors,
    output_name = "anl",
    join_fun = "dplyr::left_join"
  )

  output$mapped <- renderPrint({
    yaml::as.yaml(merged$variables())
  })

  output$merged <- renderTable({
    merged$data()[["anl"]]
  })
}

if (interactive()) {
  shinyApp(ui, server)
}
```

## Notes

- **`merge_srv()`** expects **`selectors`** to be a **named list of reactives** (as returned by **`picks_srv()`** for each selector).
- Define **`join_keys()`** on your **`teal_data`** before merging across datasets. One relationship between two datasets is enough: **`join_keys(join_key("ADSL", "ADLB", keys = "USUBJID"))`** is expanded by **`teal.data`** into a symmetric map so both names exist. Extra **`join_key("DS", "DS", …)`** self-keys are optional; they record **primary-key / row grain** (for example **`USUBJID` + `PARAM`** on long lab rows), which matters in full CDISC-style setups more than in this minimal example.
- For bookmarking, **`picks_srv()`** stores resolved picks when **`enableBookmarking = "server"`** is used on **`shinyApp()`**.
- **`values()`** filters the column(s) chosen in **`variables()`**. If **`multiple = TRUE`** variables are selected, values are derived from a **combined** representation of those columns—so do not pair **PARAM-only** level choices with a selection that also includes **`AVAL`**. Use **`values()`** with a **single** categorical column, or omit **`values()`** when taking several columns (as in this example).

