---
title: "Choices and selections in teal.picks"
author: "NEST CoreDev"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Choices and selections in teal.picks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  rmarkdown::html_vignette:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = NA)
```

## Overview

Every `picks()` object is built from three kinds of building blocks: `datasets()`, `variables()`,
and `values()`. Each one carries two core arguments:

- `choices` — the full set of options the app user can pick from.
- `selected` — the subset that is pre-selected when the app starts.

```
picks(
  datasets(choices = <what datasets the user may choose>, selected = <default dataset>),
  variables(choices = <what columns the user may choose>, selected = <default columns>),
  values(choices   = <what values the user may filter to>, selected = <default filter values>)
)
```

Choices and selections can be expressed in three ways:

| Approach | Works with |
|---|---|
| Explicit — character vector, integer index, or numeric range | `datasets()`, `variables()`, `values()` |
| `tidyselect` helpers — `everything()`, `starts_with()`, `where()`, … | `datasets()`, `variables()` only |
| Function — an R function applied at runtime to the data | `datasets()`, `variables()`, `values()` |

The sections below walk through each approach with runnable examples.

```{r data, message = FALSE}
library(teal.data)
library(teal.picks)

data <- teal_data()
data <- within(data, {
  ADSL <- data.frame(
    USUBJID  = sprintf("S%03d", 1:10),
    AGE      = sample(40:75, 10, replace = TRUE),
    SEX      = rep(c("M", "F"), 5),
    ARM      = rep(c("A", "B"), 5),
    stringsAsFactors = FALSE
  )
  ADLB <- data.frame(
    USUBJID = rep(sprintf("S%03d", 1:10), each = 3),
    PARAM   = rep(c("ALT", "AST", "CRP"), 10),
    AVAL    = round(rnorm(30, mean = 40, sd = 8), 1),
    stringsAsFactors = FALSE
  )
})

join_keys(data) <- join_keys(
  teal.data::join_key("ADSL", "ADLB", keys = "USUBJID")
)
```

---

## 1. Explicit choices

Pass a character vector to `choices` to enumerate options exactly. Use `selected` to set the
default. Integer indices work too (for example `selected = 1L` means the first element of
`choices`).

```{r explicit}
# Datasets — user may switch between ADSL and ADLB; ADSL is the default
p_datasets <- picks(
  datasets(
    choices  = c("ADSL", "ADLB"),
    selected = "ADSL"
  )
)

# Variables — only a named subset is offered; first column pre-selected
p_variables <- picks(
  datasets(choices = "ADSL", selected = "ADSL"),
  variables(
    choices  = c("AGE", "SEX", "ARM"),
    selected = "AGE",
    multiple = FALSE
  )
)

# Values — categorical filter; two levels pre-selected
p_values <- picks(
  datasets(choices = "ADSL", selected = "ADSL"),
  variables(choices = "SEX", selected = "SEX", multiple = FALSE),
  values(
    choices  = c("M", "F"),
    selected = "F"
  )
)

resolver(x = p_datasets, data = as.list(data))
resolver(x = p_variables, data = as.list(data))
resolver(x = p_values, data = as.list(data))
```

---

## 2. `tidyselect` helpers

`tidyselect` predicates let `choices` and `selected` adapt to the actual data at runtime instead
of being hard-coded. This is supported for `datasets()` and `variables()`.

Commonly used helpers:

- `tidyselect::everything()` — all items
- `tidyselect::starts_with()` / `tidyselect::ends_with()` / `tidyselect::contains()` — pattern matching
- `tidyselect::matches()` — regex matching
- `tidyselect::where(predicate)` — columns satisfying a predicate function
- `tidyselect::all_of()` / `tidyselect::any_of()` — vector-based selection (error-safe or silent)
- Integer indices such as `1L`, `1L:3L` — select by position

> Note: `tidyselect` is not supported by `values()`. Use explicit vectors or a function
> there instead (see [Section 3](#functions)).

```{r tidyselect}
# Datasets — offer any data.frame in the teal_data object
p_any_dataset <- picks(
  datasets(
    choices  = tidyselect::where(is.data.frame),
    selected = 1L  # first dataset by default
  )
)

# Variables — all numeric columns; first one pre-selected
p_numeric_vars <- picks(
  datasets(choices = "ADSL", selected = "ADSL"),
  variables(
    choices  = tidyselect::where(is.numeric),
    selected = 1L,
    multiple = FALSE
  )
)

# Variables — columns whose names start with "A"; first two pre-selected
p_a_prefix <- picks(
  datasets(choices = "ADSL", selected = "ADSL"),
  variables(
    choices  = tidyselect::starts_with("A"),
    selected = 1L:2L,
    multiple = TRUE
  )
)

resolver(x = p_any_dataset, data = as.list(data))
resolver(x = p_numeric_vars, data = as.list(data))
resolver(x = p_a_prefix, data = as.list(data))
```

---

## 3. Functions {#functions}

You can pass a plain R function as `choices` or `selected`. The function receives the relevant
context object (the current dataset for `datasets()` and `variables()`, the current column vector
for `values()`) and must return the subset to use. This is the only runtime-dynamic approach
supported by `values()`.

```{r functions}
# Variables — use the package helper is_categorical() as a column predicate.
# Without "des-delayed", the resolver calls it via vapply(data, fn, logical(1)),
# so it must accept one column and return a single logical value — which is_categorical() does.
p_cat_vars <- picks(
  datasets(choices = "ADSL", selected = "ADSL"),
  variables(
    choices  = is_categorical(),
    selected = 1L,
    multiple = TRUE
  )
)

# Values — select only even ages from the AGE column.
# Functions passed to values() must carry the "des-delayed" class so the resolver
# calls them with the column vector rather than treating them as a column predicate.
even_vals <- function(x) sort(unique(x[x %% 2 == 0]))
class(even_vals) <- append(class(even_vals), "des-delayed")

p_even_ages <- picks(
  datasets(choices = "ADSL", selected = "ADSL"),
  variables(choices = "AGE", selected = "AGE", multiple = FALSE),
  values(
    choices  = even_vals,
    selected = even_vals
  )
)

resolver(x = p_cat_vars,   data = as.list(data))
resolver(x = p_even_ages,  data = as.list(data))
```

---

## Summary

| | Explicit | `tidyselect` | Function |
|---|:---:|:---:|:---:|
| `datasets()` | yes | yes | yes |
| `variables()` | yes | yes | yes |
| `values()`    | yes | **no** | yes |

Use **explicit** choices when the set of options is fixed and known at app-development time.
Use **`tidyselect`** when you want the choices to adapt to the shape of the data without writing
a custom function. Use **functions** when the logic is more involved, or when you need runtime
behaviour for `values()`.

## See also

- `?picks` — full reference for `picks()`, `datasets()`, `variables()`, and `values()`.
- `?tidyselectors` — package-specific tidyselect helpers such as `is_categorical()`.
- `vignette("teal-picks-in-teal", package = "teal.picks")` — wiring `picks` into a `teal` module.
- `vignette("teal-picks-standalone-shiny", package = "teal.picks")` — using `picks` in plain Shiny.
