---
title: "Using teal.picks in a teal application"
author: "NEST CoreDev"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Using teal.picks in a teal application}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  rmarkdown::html_vignette:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = NA)
```

## What this package is for

`teal.picks` helps you expose analysis choices in a `teal` Shiny app in a way that mimicks how people think about data: first *which dataset*, then *which columns*, and optionally *which values* to filter to. You set all those steps at once with `picks()`; the app renders the right controls and keeps selections consistent when upstream choices change.

Typical workflow:

1. Prepare a `teal.data::teal_data()` object (when you relate multiple tables use `join_keys` so teal knows how they link).
2. Define one or more `picks()` specs: `datasets()`, optionally `variables()`, optionally `values()` for filters or ranges.
3. In your teal module, wire UI and server so those specs drive your tables, plots, or merged analysis data (see [Writing your own module](#writing-your-own-module)).

The examples below show the usual building blocks—not every app needs every level.

```{r data, message = FALSE}
library(teal)
library(teal.data)
library(teal.picks)

data <- teal_data()
data <- within(data, {
  ADSL <- teal.data::rADSL
  ADLB <- teal.data::rADLB
})

join_keys(data) <- teal.data::default_cdisc_join_keys[c("ADSL", "ADLB")]
```

### Choose a dataset

When the analysis can draw from more than one table, let the user pick the active source (here, demographics vs labs).

```{r picks-datasets}
picks_datasets <- list(
  source = picks(
    datasets(
      choices = c("ADSL", "ADLB"),
      selected = "ADLB"
    )
  )
)
```

### Choose dataset and columns

Add `variables()` so, after a dataset is chosen, the user picks one or more columns. You can list names explicitly or use tidyselect-style expressions; see `?picks` for details.

```{r picks-datasets-variables}
picks_datasets_variables <- list(
  adsl_cols = picks(
    datasets(choices = "ADSL", selected = "ADSL"),
    variables(
      choices = c("USUBJID", "AGE", "SEX"),
      selected = "AGE",
      multiple = FALSE
    )
  )
)
```

### Add a value filter (levels or ranges)

`values()` sits after `variables()`. It adapts to the column type—for example category levels for a `PARAM`-style variable, or a numeric or date range for continuous data.

<!-- # nolint start: object_length_linter. -->

```{r picks-datasets-variables-values}
picks_datasets_variables_values <- list(
  labs = picks(
    datasets(choices = "ADLB", selected = "ADLB"),
    variables(choices = "PARAM", selected = "PARAM", multiple = FALSE),
    values(
      choices = c("ALT", "AST", "CRP", "GLU"),
      selected = c("ALT", "AST"),
      multiple = TRUE
    )
  )
)
```

<!-- # nolint end: object_length_linter. -->

### Define choices and selections

Each `picks()` element has `choices` and `selected`. You can define them in four ways:

| Approach | Works with |
|---|---|
| Default behavior - no arguments needed for generic usage | `datasets()`, `variables()`, `values()`
| Static — character vector, integer index, or numeric range | `datasets()`, `variables()`, `values()` |
| `tidyselect` helpers — `everything()`, `starts_with()`, `where()`, … | `datasets()`, `variables()` only |
| Function — an R function applied at runtime to the data | `datasets()`, `variables()`, `values()` |

The sections below walk through each approach with runnable examples.

#### Defaults
Defaults depend on the slot type.
For `datasets()`, the default `choices` are all available datasets. A `picks()` call needs a `datasets()` as first positional argument, except when `check_no_dataset = FALSE`.
For `variables()`, the default `choices` is all variables in the selected dataset.
For `values()`, the default `choices` is all values of the selected variable.
For `selected`, the default is the first available choice, or all choices when `multiple = TRUE`.

```{r defaults}
picks(
  datasets(choices = "ADSL", selected = "ADSL"),
  variables()
)

picks(
  datasets(choices = "ADSL", selected = "ADSL"),
  variables(choices = "SEX", selected = "SEX", multiple = FALSE),
  values()
)
```

#### Static choices

Pass a character vector to `choices` to enumerate options exactly. Use `selected` to set the
default. Integer indices work too (for example `selected = 1L` means the first element of
`choices`).

```{r static}
# Datasets — user may switch between ADSL and ADLB; ADSL is the default
p_datasets <- picks(
  datasets(
    choices  = c("ADSL", "ADLB"),
    selected = "ADSL"
  )
)

# Variables — only a named subset is offered; first column pre-selected
p_variables <- picks(
  datasets(choices = "ADSL", selected = "ADSL"),
  variables(
    choices  = c("AGE", "SEX", "ARM"),
    selected = "AGE",
    multiple = FALSE
  )
)

# Values — categorical filter; two levels pre-selected
p_values <- picks(
  datasets(choices = "ADSL", selected = "ADSL"),
  variables(choices = "SEX", selected = "SEX", multiple = FALSE),
  values(
    choices  = c("M", "F"),
    selected = "F"
  )
)

p_datasets
p_variables
p_values
```

#### `tidyselect` helpers

`tidyselect` predicates let `choices` and `selected` adapt to the actual data at runtime instead
of being hard-coded. This is supported for `datasets()` and `variables()`.

Commonly used helpers:

- `tidyselect::everything()` — all items
- `tidyselect::starts_with()` / `tidyselect::ends_with()` / `tidyselect::contains()` — pattern matching
- `tidyselect::matches()` — regex matching
- `tidyselect::where(predicate)` — columns satisfying a predicate function
- `tidyselect::all_of()` / `tidyselect::any_of()` — vector-based selection (error-safe or silent)
- Integer indices such as `1L`, `1L:3L` — select by position

> Note: `tidyselect` is not supported by `values()`. Use explicit vectors or a function
> there instead (see [Functions](#functions)).

```{r tidyselect}
# Datasets — offer any data.frame in the teal_data object
p_any_dataset <- picks(
  datasets(
    choices  = tidyselect::where(is.data.frame),
    selected = 1L # first dataset by default
  )
)

# Variables — all numeric columns; first one pre-selected
p_numeric_vars <- picks(
  datasets(choices = "ADSL", selected = "ADSL"),
  variables(
    choices  = tidyselect::where(is.numeric),
    selected = 1L,
    multiple = FALSE
  )
)

# Variables — columns whose names start with "A"; first two pre-selected
p_a_prefix <- picks(
  datasets(choices = "ADSL", selected = "ADSL"),
  variables(
    choices  = tidyselect::starts_with("A"),
    selected = 1L:2L,
    multiple = TRUE
  )
)

p_any_dataset
p_numeric_vars
p_a_prefix
```

#### Functions {#functions}

You can pass a plain R function as `choices` or `selected`. The function receives the relevant
context object (the current dataset for `datasets()` and `variables()`, the current column vector
for `values()`) and must return the subset to use. This is the only runtime-dynamic approach
supported by `values()`.

```{r functions}
# Variables — use the package helper is_categorical() as a column predicate.
# Without "des-delayed", the resolver calls it via vapply(data, fn, logical(1)),
# so it must accept one column and return a single logical value — which is_categorical() does.
picks(
  datasets(choices = "ADSL", selected = "ADSL"),
  variables(
    choices  = is_categorical(),
    selected = 1L,
    multiple = TRUE
  )
)

# Values — select only even ages from the AGE column.
# Functions passed to values() must carry the "des-delayed" class so the resolver
# calls them with the column vector rather than treating them as a column predicate.
even_vals <- function(x) sort(unique(x[x %% 2 == 0]))
class(even_vals) <- append(class(even_vals), "des-delayed")

p_even_ages <- picks(
  datasets(choices = "ADSL", selected = "ADSL"),
  variables(choices = "AGE", selected = "AGE", multiple = FALSE),
  values(
    choices  = even_vals,
    selected = even_vals
  )
)

p_even_ages
```

#### Use of multiple to select more than one columns at once

Use `multiple = TRUE` when analysts should pass more than one variable into the next step (for example stratifiers or outcomes together).

```{r picks-multiple-variables}
picks_multiple_variables <- list(
  demo = picks(
    datasets(choices = "ADSL", selected = "ADSL"),
    variables(
      choices = c("USUBJID", "AGE", "SEX"),
      selected = c("AGE", "SEX"),
      multiple = TRUE,
      ordered = TRUE
    )
  )
)
```
See `?picks` for other arguments that affect how choices/selections are presented to users regardless of 
using static, tidyselect or functions to define them.

#### Choices and Selection Summary

| | Static | `tidyselect` | Function |
|---|:---:|:---:|:---:|
| `datasets()` | yes | yes | yes |
| `variables()` | yes | yes | yes |
| `values()`    | yes | **no** | yes |

Use **static** choices when the set of options is fixed and known at app-development time.
Use **`tidyselect`** when you want the choices to adapt to the shape of the data without writing
a custom function. Use **functions** when the logic is more involved, or when you need runtime
behavior for `values()`.


## Trying these patterns inside teal

You can paste the `picks` objects above into any module that consumes them. To quickly explore picks object(s) in teal apps, the package includes `tm_merge()`: a small example teal module that connects `picks` to a merged preview table. That is not required for normal use—it is simply a convenient host while you learn the `picks()` API.

Run the next block to explore a teal app with different picks patterns

```{r teal-app, eval = FALSE}
library(shiny)

app <- init(
  data = data,
  modules = modules(
    modules(
      label = "teal.picks patterns",
      tm_merge(
        label = "1. Dataset choice",
        picks = picks_datasets
      ),
      tm_merge(
        label = "2. Dataset & variables",
        picks = picks_datasets_variables
      ),
      tm_merge(
        label = "3. Dataset, variables & values",
        picks = picks_datasets_variables_values
      ),
      tm_merge(
        label = "4. Multiple variables",
        picks = picks_multiple_variables
      )
    )
  )
)

if (interactive()) {
  shinyApp(app$ui, app$server)
}
```

## Writing your own module {#writing-your-own-module}

In real apps you usually integrate `picks` with your own `teal::module()`. Two functions do the wiring:

- `picks_ui()` — place the selector controls in your picks object in your module UI
- `picks_srv()` — in the module server, pass `data` as a reactive `teal_data` object; it returns the resolved picks (what the user chose, after any dynamic choices are applied).


```{r module-skeleton, eval = FALSE}
tm_picks_preview <- function(label = "Custom picks module", picks) {
  teal::module(
    label = label,
    ui = function(id, picks) {
      ns <- shiny::NS(id)
      shiny::tagList(
        teal.picks::picks_ui(ns("sel"), picks = picks),
        shiny::tags$h5("Preview (first rows)"),
        shiny::tableOutput(ns("preview")),
        shiny::tags$h5("Resolved picks"),
        shiny::verbatimTextOutput(ns("resolved"))
      )
    },
    server = function(id, data, picks) {
      shiny::moduleServer(id, function(input, output, session) {
        resolved <- teal.picks::picks_srv("sel", picks = picks, data = data)
        preview_tbl <- shiny::reactive({
          shiny::req(data(), resolved())
          ds <- resolved()$datasets$selected
          vars <- resolved()$variables$selected
          shiny::req(length(ds) == 1L, length(vars) >= 1L)
          data()[[ds]][, vars, drop = FALSE]
        })
        output$preview <- shiny::renderTable({
          utils::head(preview_tbl(), 8L)
        })
        output$resolved <- shiny::renderPrint({
          shiny::req(resolved())
          str(resolved(), max.level = 2L, give.attr = FALSE)
        })
      })
    },
    ui_args = list(picks = picks),
    server_args = list(picks = picks),
    datanames = "ADSL"
  )
}

app <- init(
  data = data,
  modules = modules(
    tm_picks_preview(
      label = "Custom picks module",
      picks = picks_datasets_variables$adsl_cols
    )
  )
)

if (interactive()) {
  shinyApp(app$ui, app$server)
}
```

See `?picks_ui` and `?picks_srv` for container options and multiple named selectors.

## See also

- `?picks` — full `datasets()` / `variables()` / `values()` reference.
- `vignette("teal-picks-standalone-shiny", package = "teal.picks")` — `picks_ui()` / `picks_srv()` in plain Shiny (no `init()`).
- `teal` documentation for `init()`, `modules()`, and `teal::module()`.
